BIOS 560R Final Project
========================================================
#### Xinyi Zhao, MPH Candidate


```{r, echo=FALSE, results="hide", message=FALSE, warning=FALSE, error=FALSE}
# Set the working directory
#setwd("H:/Spring 2015/R/project")

# Load the package
library(ggplot2)
library(knitr)

# Read the data
proj <- read.csv("project560Rdata.csv", stringsAsFactors = FALSE)

# Obtain the sample size
nobs <- length(proj$id)
```
## 1. Exploratory Data Analysis

### 1.1 Descriptive Statistics

```{r, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
## Obtain frequency and percentage for dichotomous variables
# CHD
fq_chd <- as.vector(table(proj$chd))
tb_chd <- as.vector(round(prop.table(table(proj$chd))*100, digits=2))
# CAT
fq_cat <- as.vector(table(proj$cat))
tb_cat <- as.vector(round(prop.table(table(proj$cat))*100, digits=2))
# SMK
fq_smk <- as.vector(table(proj$smk))
tb_smk <- as.vector(round(prop.table(table(proj$smk))*100, digits=2))
# ECG
fq_ecg <- as.vector(table(proj$ecg))
tb_ecg <- as.vector(round(prop.table(table(proj$ecg))*100, digits=2))
# HPT
fq_hpt <- as.vector(table(proj$hpt))
tb_hpt <- as.vector(round(prop.table(table(proj$hpt))*100, digits=2))

## Obtain mean and std for continuous variables
# AGE
m_age <- format(round(c(mean(proj$age, na.rm = TRUE)), digits=2), nsmall=2)
sd_age <- format(round(c(sd(proj$age, na.rm = TRUE)), digits=2), nsmall=2)
# CHL
m_chl <- format(round(c(mean(proj$chl, na.rm = TRUE)), digits=2), nsmall=2)
sd_chl <- format(round(c(sd(proj$chl, na.rm = TRUE)), digits=2), nsmall=2)
# DBP
m_dbp <- format(round(c(mean(proj$dbp, na.rm = TRUE)), digits=2), nsmall=2)
sd_dbp <- format(round(c(sd(proj$dbp, na.rm = TRUE)), digits=2), nsmall=2)
# SBP
m_sbp <- format(round(c(mean(proj$sbp, na.rm = TRUE)), digits=2), nsmall=2)
sd_sbp <- format(round(c(sd(proj$sbp, na.rm = TRUE)), digits=2), nsmall=2)

## Create a summary table for all variables
# Create a matrix with column names
sumstat <- matrix(nrow = 9, ncol = 5)
colnames(sumstat) <- c("Characteristic", "Frequency", "%", "Mean", "SD")

# Fill in relevent information for each column
sumstat[, 1] <- c("Age (years)", "Cholesterol (mg/dL)", "Diastolic blood pressure (mmHg)", 
                  "Systolic blood pressure (mmHg)", "Coronary heart disease", 
                  "Normal catecholamine level", "Ever smoked", 
                  "Electrocardiogram abnormality", "High blood pressure")
sumstat[, 2] <- c("        ", "        ", "        ", "        ", 
                  fq_chd[2], fq_cat[2], fq_smk[2], fq_ecg[2], fq_hpt[2])
sumstat[, 3] <- c("        ", "        ", "        ", "        ", 
                  tb_chd[2], tb_cat[2], tb_smk[2], tb_ecg[2], tb_hpt[2])
sumstat[, 4] <- c(m_age, m_chl, m_dbp, m_sbp, 
                  "        ", "        ", "        ", "        ", "        ")
sumstat[, 5] <- c(sd_age, sd_chl, sd_dbp, sd_sbp, 
                  "        ", "        ", "        ", "        ", "        ")
# Display the table
kable(sumstat, align=c("l", "c", "c", "r", "r"), caption="Table 1. Characteristics for participants in the prospective study of white males in Georgia")
```


### 1.2 Exploratory Graphics
```{r, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Figure 1. Association between cholesterol (mg/dL) and age in years"}
# Create a scatter plot for the main exposure and outcome
sct <- ggplot(proj, aes(x=age, y=chl)) + geom_point(size=3.5, shape=20) + 
       ylab("Cholesterol (mg/dL)") + xlab("Age (years)") + stat_smooth() +
       ggtitle("Figure 1. Association between cholesterol (mg/dL) and age in years")
sct
```



## 2. Univariate Data Analysis

```{r, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
## Build simple logistic models for all potential predictors
## and extract odds ratio and confidence intervals
# AGE
u_age <- glm(chd ~ age, data = proj, family = "binomial")
OR_u_age <- format(round(exp(u_age$coef[2]), digits = 2), nsmall=2)
CI_u_age <- confint(u_age)
OR_CI_u_age <- format(round(exp(CI_u_age[2, ]), digits = 2), nsmall=2)
# CHL
u_chl <- glm(chd ~ chl, data = proj, family = "binomial")
OR_u_chl <- format(round(exp(u_chl$coef[2]), digits = 2), nsmall=2)
CI_u_chl <- confint(u_chl)
OR_CI_u_chl <- format(round(exp(CI_u_chl[2, ]), digits = 2), nsmall=2)
# DPB
u_dbp <- glm(chd ~ dbp, data = proj, family = "binomial")
OR_u_dbp <- format(round(exp(u_dbp$coef[2]), digits = 2), nsmall=2)
CI_u_dbp <- confint(u_dbp)
OR_CI_u_dbp <- format(round(exp(CI_u_dbp[2, ]), digits = 2), nsmall=2)
# SBP
u_sbp <- glm(chd ~ sbp, data = proj, family = "binomial")
OR_u_sbp <- format(round(exp(u_sbp$coef[2]), digits = 2), nsmall=2)
CI_u_sbp <- confint(u_sbp)
OR_CI_u_sbp <- format(round(exp(CI_u_sbp[2, ]), digits = 2), nsmall=2)
# CAT
u_cat <- glm(chd ~ cat, data = proj, family = "binomial")
OR_u_cat <- format(round(exp(u_cat$coef[2]), digits = 2), nsmall=2)
CI_u_cat <- confint(u_cat)
OR_CI_u_cat <- format(round(exp(CI_u_cat[2, ]), digits = 2), nsmall=2)
# SMK
u_smk <- glm(chd ~ smk, data = proj, family = "binomial")
OR_u_smk <- format(round(exp(u_smk$coef[2]), digits = 2), nsmall=2)
CI_u_smk <- confint(u_smk)
OR_CI_u_smk <- format(round(exp(CI_u_smk[2, ]), digits = 2), nsmall=2)
# ECG
u_ecg <- glm(chd ~ ecg, data = proj, family = "binomial")
OR_u_ecg <- format(round(exp(u_ecg$coef[2]), digits = 2), nsmall=2)
CI_u_ecg <- confint(u_ecg)
OR_CI_u_ecg <- format(round(exp(CI_u_ecg[2, ]), digits = 2), nsmall=2)
# HPT
u_hpt <- glm(chd ~ hpt, data = proj, family = "binomial")
OR_u_hpt <- format(round(exp(u_hpt$coef[2]), digits = 2), nsmall=2)
CI_u_hpt <- confint(u_hpt)
OR_CI_u_hpt <- format(round(exp(CI_u_hpt[2, ]), digits = 2), nsmall=2)

## Create a table for univariate regression results
# Create a matrix with column names
uni <- matrix(nrow = 8, ncol = 4)
colnames(uni) <- c("Predictor", "Odds Ratio", "Lower 95% CI Limit", "Upper 95% CI Limit")

# Fill in relevent information for each column
uni[, 1] <- c("Age (years)", "Cholesterol (mg/dL)", "Diastolic blood pressure (mmHg)", 
              "Systolic blood pressure (mmHg)", "Normal catecholamine level", "Ever smoked", 
              "Electrocardiogram abnormality", "High blood pressure")
uni[, 2] <- c(OR_u_age, OR_u_chl, OR_u_dbp, OR_u_sbp, OR_u_cat, OR_u_smk, OR_u_ecg, OR_u_hpt)
uni[, 3] <- c(OR_CI_u_age[1], OR_CI_u_chl[1], OR_CI_u_dbp[1], OR_CI_u_sbp[1], OR_CI_u_cat[1],
              OR_CI_u_smk[1], OR_CI_u_ecg[1], OR_CI_u_hpt[1])
uni[, 4] <- c(OR_CI_u_age[2], OR_CI_u_chl[2], OR_CI_u_dbp[2], OR_CI_u_sbp[2], OR_CI_u_cat[2],
              OR_CI_u_smk[2], OR_CI_u_ecg[2], OR_CI_u_hpt[2])

# Display the table
kable(uni, align=c("l", "c", "c", "c"), caption="Table 2. Bivariate associations between coronary heart disease and participants characteristics")
```


## 3. Multivariate Data Analysis

### 3.1 Multivariate Regression Parameters

```{r, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
## Build a multiple logistic model for SMK, CAT, ECG and HPT
## and extract odds ratio and confidence intervals
m <- glm(chd ~ smk + cat + ecg + hpt, data = proj, family = "binomial")
OR_m_CI <- exp(confint(m))
# SMK
OR_m_smk <- format(round(exp(m$coef[2]), digits = 2), nsmall=2)
OR_CI_m_smk <- format(round(OR_m_CI[2, ], digits = 2), nsmall=2)
# CAT
OR_m_cat <- format(round(exp(m$coef[3]), digits = 2), nsmall=2)
OR_CI_m_cat <- format(round(OR_m_CI[3, ], digits = 2), nsmall=2)
# ECG
OR_m_ecg <- format(round(exp(m$coef[4]), digits = 2), nsmall=2)
OR_CI_m_ecg <- format(round(OR_m_CI[4, ], digits = 2), nsmall=2)
# HPT
OR_m_hpt <- format(round(exp(m$coef[5]), digits = 2), nsmall=2)
OR_CI_m_hpt <- format(round(OR_m_CI[5, ], digits = 2), nsmall=2)

## Create a table for multivariate regression results
# Create a matrix with column names
multi <- matrix(nrow = 4, ncol = 4)
colnames(multi) <- c("Covariate", "Odds Ratio", "Lower 95% CI Limit", "Upper 95% CI Limit")

# Fill in relevent information for each column
multi[, 1] <- c("Ever smoked", "Normal catecholamine level", "Electrocardiogram abnormality", "High blood pressure")
multi[, 2] <- c(OR_m_smk, OR_m_cat, OR_m_ecg, OR_m_hpt)
multi[, 3] <- c(OR_CI_m_smk[1], OR_CI_m_cat[1], OR_CI_m_ecg[1], OR_CI_m_hpt[1])
multi[, 4] <- c(OR_CI_m_smk[2], OR_CI_m_cat[2], OR_CI_m_ecg[2], OR_CI_m_hpt[2])

# Display the table
kable(multi, align=c("l", "c", "c", "c"), caption="Table 3. Multivariate association between coronary heart disease and the set of covariates")
```

### 3.2 Multivariate Regression Graphics
```{r figcap, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap="Figure 2. Odds ratios with 95% confidence intervals for association between coronary heart disease and the set of covariates"}
# Transform the summary table into a dataframe, and rename the variables for convenience
multi_OR <- data.frame(multi)
names(multi_OR) <- c("Var", "OR", "LB", "UB")

# Transform the OR's and their CI's into numeric data
multi_OR$OR <- as.numeric(as.vector(multi_OR$OR))
multi_OR$LB <- as.numeric(as.vector(multi_OR$LB))
multi_OR$UB <- as.numeric(as.vector(multi_OR$UB))

# Create a theme for the plot             
theme_plot <- theme(axis.text.x = element_text(size=14, color="black", angle=90, hjust=1)) +
              theme(axis.text.y = element_text(size=14, color="black")) +
              theme(axis.title = element_text(size=15, color="black"))

# Plot the result of multivariate regression
plot_OR <- ggplot(multi_OR, aes(y=OR, x=Var)) + geom_point(size=3.5) +
           geom_errorbar(aes(ymin=LB, ymax=UB), width=0, size=1) + ylab("Odds ratio (95% CI)") + xlab("Covariates") +
           geom_hline(aes(yintercept=1), color="gray60",linetype="dashed", size=1) + theme_plot +
           ggtitle("Figure 2. Association between coronary heart disease and covariates")
plot_OR
```


## 4. Summary

Among `r nobs` participants in the study, average age is `r m_age` years (SD `r sd_age`); `r fq_smk[2]` (`r tb_smk[2]`%) participants have ever smoked, and `r fq_chd[2]` (`r tb_chd[2]`%) participants have been diagnosed as coronary heart disease. Based on the results of simple logistic regression, there are significant bivariate associations between coronary heart disease and age in years (OR = `r OR_u_age`; 95% CI: `r OR_CI_u_age[1]`, `r OR_CI_u_age[2]`), normal catecholamine level (OR = `r OR_u_cat`; 95% CI: `r OR_CI_u_cat[1]`, `r OR_CI_u_cat[2]`), smoking (OR = `r OR_u_smk`; 95% CI: `r OR_CI_u_smk[1]`, `r OR_CI_u_smk[2]`), electrocardiogram abnormality (OR = `r OR_u_ecg`; 95% CI: `r OR_CI_u_ecg[1]`, `r OR_CI_u_ecg[2]`) and high blood pressure (OR = `r OR_u_hpt`; 95% CI: `r OR_CI_u_hpt[1]`, `r OR_CI_u_hpt[2]`), respectively. Based on the results of multiple logistic regression, there is significant association between smoking and coronary heart disease, adjusted for normal catecholamine level, electrocardiogram abnormality and high blood pressure (adjusted OR = `r OR_m_smk`; 95% CI: `r OR_CI_m_smk[1]`, `r OR_CI_m_smk[2]`). 
